Local wind regime induced by giant linear dunes: comparison of ERA5-Land re-analysis with surface measurements

### 2022-04-02 Paper/main.tex
Adding License and readmde file
Modifying figures
Attention: Re-updating repo at first submission state (loss of data -- crash system)

\usepackage[dvipsnames]{xcolor}
\newcommand{\cg}{\textcolor{WildStrawberry}}
\newcommand{\pc}{\textcolor{ForestGreen}}
\newcommand{\new}{\textcolor{blue}}
\newcommand{\todo}{\textcolor{red}}

@@

% \title{Local wind regime induced by giant linear dunes}
\title{Local wind regime induced by giant linear dunes: comparison of  ERA5 re-analysis with surface measurements}

@@

\institute{
C. Gadal \at
Institut de M\'ecanique des Fluides de Toulouse, Universit\'e de Toulouse Paul Sabatier, CNRS, Toulouse INP-ENSEEIHT, Toulouse, France. \email{cyril.gadal@imft.fr}
\and
P. Delorme \at
School of Geography and Environmental Science, University of Southampton, Southampton, UK.
\and
C. Narteau \at
Institut de Physique du Globe de Paris, Universit\'e de Paris, CNRS, Paris, France.
\and
G. Wiggs \at
School of Geography and the Environment, University of Oxford, Oxford, UK.
\and
M. Baddock \at
Geography and Environment, Loughborough University, Loughborough, UK.
\and
J.M. Nield \at
School of Geography and Environmental Science, University of Southampton, Southampton, UK.
\and
P. Claudin \at
Physique et M\'ecanique des Milieux H\'et\'erog\`enes, CNRS, ESPCI Paris, PSL Research University, Universit\'e de Paris, Sorbonne Universit\'e, Paris, France.  \email{philippe.claudin@espci.fr}
}

@@

The description of turbulent flows over a complex topography is a rich and active subject, that is relevant for a large variety of different environmental systems \citep{Finnigan2020}. For example, the flow over hills is of primary interest for wind power, meteorological and air pollution phenomena \citep{Taylor1987}. The properties of these flows are also key to the understanding of the formation of ocean surface wind-driven waves \citep{Sullivan2010}, dissolution bedforms \citep{Claudin2017}, or sedimentary ripples and dunes \citep{Charru2013, Courrech2015}. Importantly, the troposphere presents a vertical structure, with a lower convective boundary layer, of typical kilometer-scale thickness, capped by a stably stratified region \citep{Stull1988}. The largest topographic obstacles, such as mountains can therefore interact with this upper region, and lead to internal wave generation or significant wind disturbances, such as lee-side downslope winds \citep{Durran1990}.

Focusing on the wind close to the surface, two related topographic feedbacks on the windflow over dunes can be commented on separately. First is the effect on wind speed with documented flow acceleration on upwind slopes \citep{Weaver2011} and deceleration on downwind slopes \citep{Baddock2007}, with a speed-up factor essentially proportional to the obstacle aspect ratio \citep{Jackson1975}. Importantly, the velocity maximum is typically shifted upwind of the obstacle crest. This behaviour has been theoretically predicted by means of asymptotic analysis of a neutrally stratified boundary-layer flow over an obstacle of vanishing aspect ratio \citep{Jackson1975, Mason1979, Sykes1980, Hunt1988, Belcher1998}. Experiments in flumes \citep{Zilker1977, Zilker1979, Frederick1988, Poggi2007}, in wind tunnels \citep{Gong1989, Finnigan1990, Gong1996} and in field conditions \citep{Taylor1987a, Claudin2013, Fernando2019, Lu2021, Bristow2022}, have also documented this effect. Interestingly, a similar behaviour exists for the pressure perturbation, but with a slight downwind shift for the pressure minimum \citep{Claudin2021}. The second effect, much less studied, is the flow deflection that occurs when the incident wind direction is not perpendicular to the ridge crest. While predicted to be small (less than $10^{\circ}$) in the linear regime valid for shallow topography \citep{Gadal2019}, significant flow steering has been reported in the field on the downwind side of steep enough obstacles, such as mountain ranges \citep{Kim2000, Lewis2008, Fernando2019}, well-developed sand dunes \citep{Walker2009, Hesp2015, Walker2017, Smith2017}, and valley topographies \citep{Wiggs2002}.

For practical reasons, wind measurement over sand dunes has been performed over rather small bedforms, typically a few meters high only (corresponding to tens of meters long) \citep{Wiggs1996}. Giant dunes, with kilometer-scale wavelengths and heights of tens of meters, are more difficult to investigate although for several reasons they provide a choice configuration for the study of turbulent flows over a complex topography. First, one expects larger wind disturbances for larger obstacles. Secondly, their large size makes them interact with the vertical structure of the atmosphere \citep{Andreotti2009}. Third, they usually form large patterns in sand seas and thus behave as rather clean periodic perturbations, in contrast with isolated dunes. Finally, because the morphodynamics of aeolian bedforms are strongly dependent on the local wind regime \citep{Livingstone2019}, one can expect to see the consequences of windflow disturbance by large dunes on neighbouring small dunes, in a similar manner to that reported for the effect of dunes on impact ripples \citep{Howard1977,Hood2021}.

Atmospheric flows have been much studied at the desert-scale with climate reanalyses based on global atmospheric models \citep{Blumberg1996, Livingstone2010, Ashkenazy2012, Jolivet2021, Hu2021}, such as ERA-40, ERA-Interim or ERA-5 \citep{Uppala2005, Dee2011, Hersbach2020}. However, the spatial resolution (tens of kilometers) of these reanalyses implies average quantities that do not resolve the smaller scales, ranging from individual dunes to the border of mountains \citep{Livingstone2010}. Recently, the release of ERA5-Land has resolved this limitation by providing up to 70 years of hourly wind predictions at a $9$~km spatial resolution \citep{munoz2021}. However, its validity remains to be studied, especially in remote desert areas where data assimilation is very low.

In this work, we compare local wind speeds and directions measured by meteorological stations at four different locations inside and north of the giant-dune field of the Namib sand sea to the regional predictions of the ERA5-Land climate reanalysis. Where the meteorological stations are surrounded by a relatively flat environment, we show that local measurements and regional predictions agree well. The agreement is also good in the interdune areas of the sand sea, except for some weak winds blowing at night, which exhibit an additional component aligned with the giant dune orientation, which are not predicted by the ERA5Land reanalysis (section 2). Further, we are able to link the magnitude of these differences to the circadian cycle of the atmospheric boundary layer (section 3). Finally, we draw implications for the wind disturbances on smaller-scale dunes (section 4), suggesting a possible origin for crossing dunes.

@@

%---------------------------------------------------------------------------------------
\section{Wind regimes across the Namib Sand Sea}
We measured the wind regime at four different locations in Namibia, representative of various arid environments across the Namib desert (Fig.~\ref{Fig1}, Online Resource Fig.~\ref{Fig1_supp}). The Adamax station was located at the Adamax waterhole to the west of Etosha Pan in northern Namibia, in a sparsely vegetated area. The Huab station was near the coast on a hyper-arid flat gravel plain lying above the ephemeral Huab river. Here, 5--6~m high barchan dunes develop from the sediment blowing out of the river valley. These two stations were located in a relatively flat environments. In contrast, the North Sand Sea and South Sand Sea stations were located in the interdunes between 50--100~m high linear dunes with kilometer-scale wavelengths and superimposed patterns. In this section, we describe and compare winds from local measurements and climate reanalysis predictions.

%------------------------
\subsection{Wind and elevation data}
At each meteorological station (Fig.~\ref{Fig1}), wind speed and direction were sampled every 10 minutes using cup anemometers and wind vanes at heights between 2~m and 3~m depending on the station. The available period of measurements at each station ranged from 1 to 5 discontinuous years distributed between 2012 and 2020 (Online Resource Fig.~\ref{Fig2_supp}). We checked that at least one complete seasonal cycle was available for each station. Regional winds were extracted at the same locations and periods from the ERA5-Land dataset, which is a replay at a smaller spatial resolution of ERA5, the latest climate reanalysis from the ECMWWF \citep{Hersbach2020, munoz2021}. This dataset provided hourly predictions of the 10-m wind velocity and direction at a spatial resolution of $0.1^\circ\times0.1^\circ$ ($\simeq 9~\textup{km}$ in Namibia).

@@

Dune properties were computed using autocorrelation on the 30-m Digital Elevation Models (DEMs) of the shuttle radar topography mission \citep{Farr2007}. For the South Sand Sea and North Sand Sea stations, we obtain, respectively, orientations of $85^\circ$ and $125^\circ$ with respect to the North, wavelengths of $2.6~\textup{km}$ and $2.3~\textup{km}$ and amplitudes (or half-heights) of $45~\textup{m}$ and $20~\textup{m}$ (Online Resource Fig.~\ref{Fig5_supp} for more details).

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure2.pdf}
\caption{Temporal comparison between the wind data coming from the ERA5-Land climate reanalysis (orange lines) and from the local measurements (blue lines). Coloured swathes indicate day (between 1000 UTC and 2200 UTC) and night (before 1000 UTC or after 2200 UTC) \textbf{a--b}: Adamax station in summer. \textbf{b--c}: Adamax station in winter. \textbf{d--e}: North Sand Sea station in summer. \textbf{f--g}: North Sand Sea station in winter. Time series of the two other stations are shown in Online Resource Fig.~\ref{Fig6_supp}.}
\label{Fig2}
\end{figure}
%---------------------------------------------

%------------------------
\subsection{Comparison of local and regional winds}
\label{section_data_feedback}
The obtained measured and predicted wind regimes are shown in figure~\ref{Fig1}. In the Namib, the regional wind patterns are essentially controlled by the see breeze, resulting in strong northward components (sometimes slightly deviated by the large scale topography) present in all regional wind roses \citep{lancaster1985}. These daytime winds are dominant during the period October-March (Fig.~\ref{Fig2}f and Online Resource Fig.~\ref{Fig6_supp}f). During April-September, an additional easterly component can be recorded during the night, induced by the combination of katabatic winds forming in the mountains, and infrequent `berg' winds, which are responsible for the high wind velocities observed \citep{lancaster1984}. The frequency of these easterly components decreases from inland to the coast. As a result, bidirectional wind regimes within the Namib Sand Sea and at the Adamax salt pan (Fig.~\ref{Fig1}b,d,e) and a unidirectional wind regime on the coast at the outlet of the Huab River (Fig.~\ref{Fig1}c) are observed.

In the case of the Adamax and Huab stations, the time series of wind speed and direction from the regional predictions quantitatively match those corresponding to the local measurements (Fig.~\ref{Fig2}a--d) and Online Resource Fig.~\ref{Fig6_supp}a--d,~\ref{Fig7_supp}). For the North Sand Sea and South Sand Sea stations within the giant dune field, we observe that this agreement is also good, but limited to the October-March time period (Fig.~\ref{Fig2}e--h and and Online Resource Fig.~\ref{Fig6_supp}e--h). As a matter of fact, the measured wind roses exhibit additional wind components aligned with the giant dune orientation, as evidenced on the satellite images (Fig.~\ref{Fig1}c,d).

More precisely, during the April-September period, the local and regional winds in the interdune match during daytime only, i.e when the southerly/southwesterly sea breeze dominates (Figs.~\ref{Fig2}e,f and \ref{Fig3}, Online Resource Fig.~\ref{Fig8_supp}). In the late afternoon and during the night, when the northwesterly `berg' and katabatic winds blow, measurements and predictions differ. In this case, the angular wind distribution of the local measurements exhibits two additional modes corresponding to reversing winds aligned with the giant dune orientation (purple frame in Fig.~\ref{Fig3}, Online Resource Figs.~\ref{Fig8_supp} and \ref{Fig9_supp}). This deviation is also associated with a general attenuation of the wind strength (Online Resource Fig.~\ref{Fig10_supp}). Remarkably, all these figures show that these wind reorientation and attenuation processes occur only at low wind velocities, typically for $u_{*} \lesssim 0.1~\textrm{m}~\textrm{s}^{-1}$. For shear velocities larger than $\simeq 0.25~\textrm{m}~\textrm{s}^{-1}$, the wind reorientation is not apparent. Finally, for intermediate shear velocities, both situations of wind flow reoriented along the dune crest and not reoriented can be successively observed (Online Resource Fig.~\ref{Fig9_supp}).

@@

%---------------------------------------------------------------------------------------
\section{Influence of wind speed and circadian cycle on the atmospheric boundary layer}
The wind deflection induced by linear dunes has previously been related to the incident angle between wind direction and crest orientation, with a maximum deflection evident for incident angles between $30^{\circ}$ and $70^{\circ}$ \citep{Walker2009, Hesp2015}. In the data analysed here, the most deflected wind at both the North and South Sand Sea stations is seen to be where the incident angle is perpendicular to the giant dunes (Figs.~\ref{Fig1} and \ref{Fig3}, Online Resource Fig.~\ref{Fig8_supp}). It therefore appears that the incident wind angle is not the dominant control on maximum wind deflection. Further, and as shown in Figure~\ref{Fig3}, winds of high and low velocities show contrasting behaviour in characteristics of deflection. This suggests a change in hydrodynamical regime between the winds. In this section, we discuss the relevant parameters associated with the dynamical mechanisms that govern the interactions between the atmospheric boundary layer flow and giant dune topographies. This analysis allows us to provide a physics-based interpretation of our measured wind data.

@@

With this theoretical framework in mind, and in the context of the measured wind data in the North and South Sand Sea stations, the smallest wind disturbances are expected to occur during the day, when the ABL depth is the largest and comparable to the dune wavelength ($k H \gtrsim 1$), which corresponds to a weak confinement situation (Fig.~\ref{Fig4}c,d). In contrast, large wind disturbances are expected to occur during the night, when the confinement is mainly induced by a shallow ABL (Fig.~\ref{Fig4}e). However, this strong confinement can be somewhat reduced in the case of strong winds, corresponding to large values of the Froude number (Fig.~\ref{Fig4}f). This is in qualitative agreement with the transition from deflected to non-deflected winds related to low and high velocities observed in our data (Sec.~\ref{section_data_feedback}).

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure5.pdf}
\caption{Regime diagrams of the wind deviation $\delta_{\theta}$ (\textbf{a}) and relative attenuation/amplification $\delta_{u}$ (\textbf{b}) in the space $(\mathcal{F}, \, kH)$, containing the data from both the North Sand Sea and South Sand Sea stations. Blue dashed lines empirically delimit the different regimes. The point density in each bin of the diagrams is shown in Online Resource Fig.~\ref{Fig13_supp} -- 95\% of the data occur in the range $-1 < \delta u < 1$. Similar regime diagrams in the spaces $(\mathcal{F}_{\textup{I}}, \, kH)$ and $(\mathcal{F}_{\textup{I}}, \, \mathcal{F})$ are shown in Online Resource Fig.~\ref{Fig14_supp}.}
\label{Fig5}
\end{figure}
%---------------------------------------------

%------------------------
\subsection{Data distribution in the flow regimes}
We can go one step further and analyse how our data quantitatively spread over the different regimes discussed above. For that purpose, one needs to compute $kH$ and $\mathcal{F}$ from the time series. $H$, $U$ and the other atmospheric parameters can be deduced from the various vertical profiles (temperature, humidity) available in the ERA5 climate reanalysis (Online Resource section 2). We quantify the flow deflection $\delta_{\theta}$ as the minimal angle between the wind orientations comparing the local measurements and the regional predictions. We also compute the relative velocity modulation as
%
\begin{equation}
\delta_{\textup{u}} = \frac{u_{*}^{\textup{ERA}} -  u_{*}^{\textup{station}}}{u_{*}^{\textup{ERA}}}.
\end{equation}
%
These two quantities are represented as maps in the plane $(\mathcal{F}, \, kH)$ (Fig.~\ref{Fig5}a,b), and one can clearly identify different regions in these graphs. Small wind disturbances (small $\delta_{\theta}$ and $\delta_{\textup{u}}$) are located in the top-right part of the diagrams, corresponding to a regime with low-interaction as well as low-confinement ($k H$ and $\mathcal{F}$ large enough, Fig.~\ref{Fig4}d). Lower values of $k H$ (stronger interaction) or of Froude number (stronger confinement) both lead to an increase in wind disturbances, both in terms of orientation and velocity. Below a crossover value $k H \simeq 0.3$, wind disturbance is less sensitive to the $\mathcal{F}$-value. This is probably due to enhanced non-linear effects linked to flow modulation by the obstacle when confinement is strong. The Froude number also controls a transition from damped to amplified wind velocities in the interdune, with a crossover around $\mathcal{F} \simeq 0.4$ (Fig.~\ref{Fig5}b). Such an amplification is rather unexpected. Checking the occurrence of the corresponding data, it appears that these amplifications are associated with the southerly sea breeze, and occur dominantly during the October-March period, when the other easterly wind is not present (Online Resource Fig.~\ref{Fig15_supp}a--b). Furthermore, they occur less frequently during the afternoon, and more frequently at the end of the day (Online Resource Fig.~\ref{Fig15_supp}c). This effect may be linked to a change in the flow behaviour in the lee side of the obstacles but further measurements are clearly needed in order to assess the different possibilities \citep{baines1995, Vosper2004}.

@@

ERA5 and ERA5-Land datasets are publicly available at the Copernicus Climate Change Service (C3S) Climate Data Store. The locally measured wind data can be found at \cg{upload on public data repository}. The digital elevation models from the Shuttle Radar Topography Mission are publicly available from Nasa servers, and can be downloaded at https://dwtkns.com/srtm30m/. Fully documented codes used to analyse this study are available at https://github.com/Cgadal/GiantDunes (will be made public upon acceptance of this manuscript for publication).

[citing all relevant grants ...] \pc{TOAD}
\end{acknowledgements}

>>>>

@@

\title{Local wind regime induced by giant linear dunes: comparison of ERA5-Land re-analysis with surface measurements}

@@

\institute{
C. Gadal \at
Institut de M\'ecanique des Fluides de Toulouse, Universit\'e de Toulouse Paul Sabatier, CNRS, Toulouse INP-ENSEEIHT, Toulouse, France. \email{cyril.gadal@imft.fr}
\and
P. Delorme \at
Energy and Environment Institute, University of Hull, Hull, UK.
\and
C. Narteau \at
Institut de Physique du Globe de Paris, Universit\'e de Paris, CNRS, Paris, France.
\and
G. Wiggs \at
School of Geography and the Environment, University of Oxford, Oxford, UK.
\and
M. Baddock \at
Geography and Environment, Loughborough University, Loughborough, UK.
\and
J.M. Nield \at
School of Geography and Environmental Science, University of Southampton, Southampton, UK.
\and
P. Claudin \at
Physique et M\'ecanique des Milieux H\'et\'erog\`enes, CNRS, ESPCI Paris, PSL Research University, Universit\'e de Paris, Sorbonne Universit\'e, Paris, France.
}

@@

The description of turbulent flows over complex topography is relevant for a large variety of different environmental systems \citep{Finnigan2020}. For example, the flow over hills is of primary interest for wind power, meteorological and air pollution phenomena \citep{Taylor1987}. The properties of these flows are also key to the understanding of geophysical phenomena, including the formation of wind-driven waves on the ocean surface \citep{Sullivan2010}, dissolution bedforms \citep{Claudin2017}, or sedimentary ripples and dunes \citep{Charru2013, Courrech2015}. Importantly, the troposphere presents a vertical structure, with a lower convective boundary layer, of typical kilometer-scale thickness, capped by a stably stratified region \citep{Stull1988}. The largest topographic obstacles, such as mountains, can therefore interact with this upper region and lead to internal wave generation or significant wind disturbances, such as lee-side downslope winds \citep{Durran1990}.

Focusing on the wind close to the surface, two related topographic feedbacks on the windflow over dunes can be commented on separately. First is the effect on wind speed, with documented flow acceleration on upwind slopes \citep{Weaver2011} and deceleration on downwind slopes \citep{Baddock2007}, where the speed-up factor is essentially proportional to the obstacle aspect ratio \citep{Jackson1975}. Importantly, the velocity maximum is typically shifted upwind of the obstacle crest. This behaviour has been theoretically predicted by means of asymptotic analysis of a neutrally stratified boundary-layer flow over an obstacle of vanishing aspect ratio \citep{Jackson1975, Mason1979, Sykes1980, Hunt1988, Belcher1998}. Experiments in flumes \citep{Zilker1977, Zilker1979, Frederick1988, Poggi2007, Bristow2022}, in wind tunnels \citep{Gong1989, Finnigan1990, Gong1996} and in field conditions \citep{Taylor1987a, Claudin2013, Fernando2019, Lu2021}, have also documented this effect. Interestingly, a similar behaviour exists for the pressure perturbation, but with a slight downwind shift for the pressure minimum \citep{Claudin2021}. The second effect, much less studied, is the flow deflection that occurs when the incident wind direction is not perpendicular to the ridge crest. While predicted to be small (less than $10^{\circ}$) in the linear regime valid for shallow topography \citep{Gadal2019}, significant flow steering has been reported in the field on the downwind side of steep enough obstacles, such as mountain ranges \citep{Kim2000, Lewis2008, Fernando2019}, well-developed sand dunes \citep{Walker2009, Hesp2015, Walker2017, Smith2017, deWinter2020}, and valley topographies \citep{Wiggs2002, Garvey2005}.

For practical reasons, wind measurement over sand dunes has been performed over small bedforms, typically a few meters high (corresponding to tens of meters long) (e.g.~\citet{Lancaster1996}, \citet{MckennaNeuman1997}, \citet{Sauermann2003}, \citet{Andreotti2002}, \citet{Walker2002}, \citet{Weaver2011}). Giant dunes, with kilometer-scale wavelengths and heights of tens of meters, are more difficult to investigate although for several reasons they provide a choice configuration for the study of turbulent flows over a complex topography. First, one expects larger wind disturbances for larger obstacles. Secondly, their large size makes them interact with the vertical structure of the atmosphere \citep{Andreotti2009}. Third, they usually form large patterns in sand seas and thus behave as rather clean periodic perturbations, in contrast with isolated dunes. Finally, because the morphodynamics of aeolian bedforms are strongly dependent on the local wind regime \citep{Livingstone2019}, one can expect to see the consequences of windflow disturbance by large dunes on neighbouring small dunes. A similar effect is observed on the properties of impact ripple patterns due to the presence of dunes \citep{Howard1977,Hood2021}.

Atmospheric flows have been much studied at the desert-scale with climate reanalyses based on global atmospheric models \citep{Blumberg1996, Livingstone2010, Ashkenazy2012, Jolivet2021, Hu2021}, such as ERA-40, ERA-Interim or ERA-5 \citep{Uppala2005, Dee2011, Hersbach2020}. However, the spatial resolution (tens of kilometers) of these reanalyses implies average quantities that do not resolve the smaller scales of interest, which range from individual dunes to small mountains \citep{Livingstone2010}. Recently, the release of ERA5-Land has resolved this limitation by providing up to 70 years of hourly wind predictions at a $9$~km spatial resolution \citep{munoz2021}. However, its validity remains to be studied, especially in remote desert areas where assimilation of measured data is very low.

In this work, we compare local wind speeds and directions measured by meteorological stations at four different locations inside and north of the giant-dune field of the Namib sand sea to the regional predictions of the ERA5-Land climate reanalysis. Where the meteorological stations are surrounded by a relatively flat environment, we show that local measurements and regional predictions agree well. The agreement is also good in the interdune areas of the sand sea, except for some weak winds blowing at night, which exhibit an additional component aligned with the giant dune orientation. These winds are not predicted by the ERA5-Land reanalysis (section 2). Further, we are able to link the magnitude of these differences to the circadian cycle of the atmospheric boundary layer (section 3). Finally, we draw implications for the wind disturbances on smaller-scale dunes (section 4), suggesting a possible origin for crossing dunes.

@@

%---------------------------------------------------------------------------------------
\section{Wind regimes across the Namib Sand Sea}
We measured the wind regime at four different locations in Namibia, representative of various arid environments across the Namib desert (Fig.~\ref{Fig1}, Online Resource Fig.~\ref{Fig1_supp}). The Etosha West station was located at the Adamax waterhole to the west of Etosha Pan in northern Namibia, in a sparsely vegetated area. The Huab station was near the coast on a hyper-arid flat gravel plain lying north the ephemeral Huab river. Here, barchan dunes up to a few meters in height develop from the sediment blowing out of the river valley \citep{Nield2017, Hesp1998}. These two stations were both located in relatively flat environments. In contrast, the North Sand Sea and South Sand Sea stations were located in the interdunes between linear dunes with kilometer-scale wavelengths, hectometer-scale heights and superimposed patterns. In this section, we describe and compare winds from local measurements and climate reanalysis predictions.

%------------------------
\subsection{Wind and elevation data}
At each meteorological station (Fig.~\ref{Fig1}), wind speed and direction were sampled every 10 minutes using cup anemometers (Vector Instruments A100-LK) and wind vanes (Vector Instruments W200-P) at heights which varied between 2~m and 3~m depending on the station. The available period of measurements at each station ranged from 1 to 5 discontinuous years distributed between 2012 and 2020 (Online Resource Fig.~\ref{Fig2_supp}). We checked that at least one complete seasonal cycle was available for each station. Regional winds were extracted at the same locations and periods from the ERA5-Land dataset, which is a replay at a smaller spatial resolution of ERA5, the latest climate reanalysis from the ECMWWF \citep{Hersbach2020, munoz2021}. This dataset provided hourly predictions of the 10-m wind velocity and direction at a spatial resolution of $0.1^\circ\times0.1^\circ$ ($\simeq 9~\textup{km}$ in Namibia).

@@

Dune properties were computed using autocorrelation on the 30-m Digital Elevation Models (DEMs) of the shuttle radar topography mission \citep{Farr2007}. For the North and South Sand Sea stations, we obtain, respectively, orientations of $85^\circ$ and $125^\circ$ with respect to the North, wavelengths of $2.6~\textup{km}$ and $2.3~\textup{km}$ and amplitudes (or half-heights) of $45~\textup{m}$ and $20~\textup{m}$ (Online Resource Fig.~\ref{Fig5_supp} for more details). This agrees with direct measurements made on site.

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure2.pdf}
\caption{Temporal comparison between the wind data coming from the ERA5-Land climate reanalysis (orange lines) and from the local measurements (blue lines). Coloured swathes indicate day (between 1000 UTC and 2200 UTC) and night (before 1000 UTC or after 2200 UTC) \textbf{a--b}: Etosha West station in summer. \textbf{b--c}: Etosha West station in winter. \textbf{d--e}: North Sand Sea station in summer. \textbf{f--g}: North Sand Sea station in winter. Time series of the two other stations are shown in Online Resource Fig.~\ref{Fig6_supp}.}
\label{Fig2}
\end{figure}
%---------------------------------------------

%------------------------
\subsection{Comparison of local and regional winds}
\label{section_data_feedback}
The measured and predicted wind regimes are shown in Fig.~\ref{Fig1}. In the Namib, the regional wind patterns are essentially controlled by the sea breeze, resulting in strong northward components (sometimes slightly deviated by the large scale topography) present in all regional wind roses \citep{lancaster1985}. These daytime winds are dominant during the period October-March (Fig.~\ref{Fig2}f and Online Resource Fig.~\ref{Fig6_supp}f). During April-September, an additional (and often nocturnal) easterly component can also be recorded, induced by the combination of katabatic winds forming in the mountains, and infrequent `berg' winds, which are responsible for the high wind velocities observed \citep{lancaster1984}. The frequency of these easterly components decreases from inland to the coast. As a result, bidirectional wind regimes within the Namib Sand Sea and at the west Etosha site (Fig.~\ref{Fig1}b,d,e) and a unidirectional wind regime on the coast at the outlet of the Huab River (Fig.~\ref{Fig1}c) are observed.

In the case of the Etosha West and Huab stations, the time series of wind speed and direction from the regional predictions quantitatively match those corresponding to the local measurements (Fig.~\ref{Fig2}a--d) and Online Resource Figs.~\ref{Fig6_supp}a--d,~\ref{Fig7_supp},~\ref{Fig9_supp}). For the North Sand Sea and South Sand Sea stations within the giant dune field, we observe that this agreement is also good, but limited to the October-March time period (Fig.~\ref{Fig2}e--h and and Online Resource Fig.~\ref{Fig6_supp}e--h). However, the field-measured wind roses exhibit additional wind components aligned with the giant dune orientation, as evidenced on the satellite images (Fig.~\ref{Fig1}c,d).

More precisely, during the April-September period, the local and regional winds in the interdune match during daytime only, i.e when the southerly/southwesterly sea breeze dominates (Figs.~\ref{Fig2}e,f and \ref{Fig3}, Online Resource Fig.~\ref{Fig8_supp}). In the late afternoon and during the night, when the easterly `berg' and katabatic winds blow, measurements and predictions differ. In this case, the angular wind distribution of the local measurements exhibits two additional modes corresponding to reversing winds aligned with the giant dune orientation (purple frame in Fig.~\ref{Fig3}, Online Resource Figs.~\ref{Fig8_supp} and \ref{Fig10_supp}). This deviation is also associated with a general attenuation of the wind strength (Online Resource Fig.~\ref{Fig11_supp}). Remarkably, all these figures show that these wind reorientation and attenuation processes occur only at low velocities of the regional wind, typically for $u_{*, \, \textup{ERA}} \lesssim 0.2~\textrm{m}~\textrm{s}^{-1}$. For shear velocities larger than $u_{*, \, \textup{ERA}} \simeq 0.3~\textrm{m}~\textrm{s}^{-1}$, the wind reorientation is not apparent. Finally, for intermediate shear velocities, both situations of wind flow reoriented along the dune crest and not reoriented can be successively observed (Online Resource Fig.~\ref{Fig10_supp}). Importantly, these values are not precise thresholds, but indicative of a crossover between regimes, whose physical interpretation is discussed in the next section.

@@

%---------------------------------------------------------------------------------------
\section{Influence of wind speed and circadian cycle on the atmospheric boundary layer}
The wind deflection induced by linear dunes has previously been related to the incident angle between wind direction and crest orientation, with a maximum deflection evident for incident angles between $30^{\circ}$ and $70^{\circ}$ \citep{Walker2009, Hesp2015}. In the data analysed here, the most deflected wind at both the North and South Sand Sea stations is seen to be where the incident angle is perpendicular to the giant dunes (Figs.~\ref{Fig1} and \ref{Fig3}, Online Resource Fig.~\ref{Fig8_supp}). It therefore appears that in our case, the incident wind angle is not the dominant control on maximum wind deflection. Further, and as shown in Fig.~\ref{Fig3}, winds of high and low velocities show contrasting behaviour in characteristics of deflection. This suggests a change in hydrodynamical regime between the winds. In this section, we discuss the relevant parameters associated with the dynamical mechanisms that govern the interactions between the atmospheric boundary layer flow and giant dune topographies. This analysis allows us to provide a physics-based interpretation of our measured wind data.

@@

With this theoretical framework in mind, and in the context of the measured wind data in the North and South Sand Sea stations, the smallest wind disturbances are expected to occur during the day, when the ABL depth is the largest and comparable to the dune wavelength ($k H \gtrsim 1$), which corresponds to a weak confinement situation (Fig.~\ref{Fig4}c,d). In contrast, large wind disturbances are expected to occur during the night, when the confinement is mainly induced by a shallow ABL (Fig.~\ref{Fig4}e). However, this strong confinement can be somewhat reduced in the case of strong winds, corresponding to large values of the Froude number and a less `rigid' interface (Fig.~\ref{Fig4}f). This is in qualitative agreement with the transition from deflected to non-deflected winds related to low and high velocities observed in our data (Sec.~\ref{section_data_feedback}).

%---------------------------------------------
\begin{figure}
\centering
\includegraphics[scale=1]{Figures/Figure5.pdf}
\caption{Regime diagrams of the wind deviation $\delta_{\theta}$ (\textbf{a}) and relative attenuation/amplification $\delta_{u}$ (\textbf{b}) in the space $(\mathcal{F}, \, kH)$, containing the data from both the North Sand Sea and South Sand Sea stations. Green dashed lines empirically delimit the different regimes. The point density in each bin of the diagrams is shown in Online Resource Fig.~\ref{Fig14_supp} -- 95\% of the data occur in the range $-1 < \delta u < 1$. Similar regime diagrams in the spaces $(\mathcal{F}_{\textup{I}}, \, kH)$ and $(\mathcal{F}_{\textup{I}}, \, \mathcal{F})$ are shown in Online Resource Fig.~\ref{Fig15_supp}.}
\label{Fig5}
\end{figure}
%---------------------------------------------

%------------------------
\subsection{Data distribution in the flow regimes}
We can go one step further and analyse how our data quantitatively spread over the different regimes discussed above. For that purpose, one needs to compute $kH$ and $\mathcal{F}$ from the time series. $H$, $U$ and the other atmospheric parameters can be deduced from the various vertical profiles (temperature, humidity) available in the ERA5 climate reanalysis (Online Resource section 2). We quantify the flow deflection $\delta_{\theta}$ as the minimal angle between the wind orientations comparing the local measurements and the regional predictions. We also compute the relative velocity modulation as
%
\begin{equation}
\delta_{\textup{u}} = \frac{u_{*}^{\textup{ERA}} -  u_{*}^{\textup{station}}}{u_{*}^{\textup{ERA}}}.
\end{equation}
%
These two quantities are represented as maps in the plane $(\mathcal{F}, \, kH)$ (Fig.~\ref{Fig5}a,b), and one can clearly identify different regions in these graphs. Small wind disturbances (small $\delta_{\theta}$ and $\delta_{\textup{u}}$) are located in the top-right part of the diagrams, corresponding to a regime with low-interaction as well as low-confinement ($k H$ and $\mathcal{F}$ large enough, Fig.~\ref{Fig4}d). Lower values of $k H$ (stronger interaction) or of Froude number (stronger confinement) both lead to an increase in wind disturbances, both in terms of orientation and velocity. Below a crossover value $k H \simeq 0.3$, wind disturbance is less sensitive to the $\mathcal{F}$-value. This is probably due to enhanced non-linear effects linked to flow modulation by the obstacle when confinement is strong. The Froude number also controls a transition from damped to amplified wind velocities in the interdune, with a crossover around $\mathcal{F} \simeq 0.4$ (Fig.~\ref{Fig5}b). Such an amplification is rather unexpected. Checking the occurrence of the corresponding data, it appears that these amplifications are associated with the southerly sea breeze, and occur dominantly during the October-March period, when the other easterly wind is not present (Online Resource Fig.~\ref{Fig16_supp}a--b). Furthermore, they occur less frequently during the afternoon, and more frequently at the end of the day (Online Resource Fig.~\ref{Fig16_supp}c). This effect may be linked to a change in the flow behaviour in the lee side of the obstacles but further measurements are needed in order to assess the different possibilities \citep{baines1995, Vosper2004}.

@@

All data used in this study can be found in \citet{DataPaper}. Note that it contains modified Copernicus Climate Change Service Information (2021). Neither the European Commission nor ECMWF is responsible for any use that may be made of the Copernicus Information or Data it
contains. Fully documented codes used to analyse this study are available at https://github.com/Cgadal/GiantDunes (will be made public upon acceptance of this manuscript for publication).

% ERA5 and ERA5-Land datasets are publicly available at the Copernicus Climate Change Service (C3S) Climate Data Store. As such, It contains modified Copernicus Climate Change
% Service Information (2021). Neither the European Commission nor ECMWF is
% responsible for any use that may be made of the Copernicus Information or Data it
% contains. The locally measured wind data can be found at \cg{upload on public data repository}. The digital elevation models from the Shuttle Radar Topography Mission are publicly available from Nasa servers, and can be downloaded at https://dwtkns.com/srtm30m/. Fully documented codes used to analyse this study are available at https://github.com/Cgadal/GiantDunes (will be made public upon acceptance of this manuscript for publication).
